An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-07 main.tex
more proof

A monoid $\Monoid = \Tuple{X;\MonoidOp;0_{\Monoid}}$ is an algebraic structure
consisting of the carrier set of elements, $X$, an associative binary operation
$X \times X \rightarrow X$ denoted as $\MonoidOp$, that is where for all $a, b,
c \in X$, $(a \MonoidOp b) \MonoidOp c = a \MonoidOp (b \MonoidOp c)$. Finally,
$\Monoid$ must have an identity element $0_{\Monoid} \in X$ such that
$0_{\Monoid} \MonoidOp a = a \MonoidOp 0_{\Monoid} =   a$ for every $a \in X$.
We sometimes use integer multiplication to represent repeated application of
$\MonoidOp$, e.g. $3a = a \MonoidOp a \MonoidOp a$, for $a \in X$. $\Monoid$ is
called \textit{commutative} if $\MonoidOp$ also commutes, that is if $a
\MonoidOp b = b \MonoidOp a$ for all $a, b \in X$.

Finally, we sometimes refer to a structure-preserving map between two monoids
$\Monoid_1, \Monoid_2$ as a \textit{homomorphism} between $\Monoid_1$ and
$\Monoid_2$, that is a map $\Map : S_1 \rightarrow S_2$ such that $\Map(a
\MonoidOp_1 b) = \Map(a) \MonoidOp_2 \Map(b)$, where $S_1, S_2$ are the carrier
sets of $\Monoid_1$ and $\Monoid_2$ respectively, and $\MonoidOp_1, \MonoidOp_2$
their binary operations.

@@

We define an alphabet as a set of symbols $\Alphabet$ with words $\Strings$, and
the concatenation operation as $s_1 \Concat{} s_2$ over two strings $s_1, s_2$.
Note that $\Strings = \Tuple{\Alphabet;\Concat{};\epsilon}$, is a
non-commutative monoid, referred to as the free monoid. The string length
function, $\Length{s}$ is an example of a homomorphism between $\Strings$
and~$\mathbb{Z}$.

@@

\subsection{The Parikh image of a regular language expressed in Presburger arithmetic}

@@

It is easy to see that the Parikh map~$\ParikhMap$ represents a
homomorphism from the (free) non-commutative monoid~$\Strings$ to the
(free) commutative monoid~$\Naturals^k$. As we are often not
interested in the full Parikh image, but only in some projections of
it, for the purposes of this paper we use a generalised version of the
Parikh map. We consider \emph{arbitrary} homomorphisms
$\Map:\: \Sigma^* \to \Monoid$, where
$\Monoid = \left(X;\MonoidOp;0_{\Monoid}\right)$ is a commutative
monoid; this means that $\Map(\epsilon) = 0_{\Monoid}$ and
$\Map(u \Concat w) = \Map(u) \MonoidOp \Map(w)$. We give several
examples of such generalised Parikh image later in this section.

@@

\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$, for some automaton
  $\Automaton = \AutomatonTuple$, modulo some map $\Map$ to a commutative monoid
  $\Monoid$ as described in \cref{sec:generalised} and with a transition
  selection function $\Filter:\: \Transitions \to \Naturals$ holds when
  $\MonoidElement \in \Monoid$ is the Parikh image of $\Automaton$ modulo
  $\Map$, or more formally when there is an accepting path $\Path =
  \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such that:
  \begin{itemize}
    \item $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
    \item $\MonoidElement = \Map(\WordOf(\Path))$
  \end{itemize}
\end{definition}

@@

Note that the filtering function~$\Filter$ is evaluated symbolically in these
rules, and can in practice be read as a symbolic function from transitions to
terms (e.g.~\texttt{t} or~\texttt{t+1}). In our implemendation \Catra{},
described in \cref{sec:implementation}, $\Filter$~is a vector of fresh variables
with the same size as~$\Transitions$.

We use the shorthand notation~$\Transitions_\Automaton$ to refer to the
transitions of an automaton~$\Automaton$. Additionally, for an
automaton~$\Automaton = \AutomatonTuple$ we allow mapping the selection function
like so: $\Filter(\Automaton) = \Tuple{\States, \InitialState, \AcceptingStates,
\Set{\Transition \in \Transitions \SuchThat \Filter(\Transition) > 0}}$, i.e.
$\Automaton$~with only the transitions for which~$\Filter$ is positive. In this
instance, the basis for the matched linear inequalities is
implicitly~$\SomeInequalities$. Similarly, for our commutative monoid $\Monoid =
\left(X;\MonoidOp;0_{\Monoid}\right)$ and the map into it $\Map : \Strings
\rightarrow X$, we also allow mapping over transitions:
$\Map(\FromLabelTo{\State}{\Label}{\State'}) = \Map(\Label)$.

The rules of \cref{tbl:rules:single} are meant to be used in tableaux-style
proofs, and are read bottom-up, where the syntax at the bottom is replaced with
the one at the top until \Fudge{it is possible to close the proof goal}. Since
our rules operate by adding and matching linear (in)equalities in a proof goal,
we use the shorthand of listing the matched inequalities as antecedents. An
example of this can be seen in the~\Propagate{} rule. \Fudge{We pull solutions
out of a hat? I have no idea.}

@@

\subsubsection{\Calculus{} computes the homomorphic image}
\Fudge{I am unsure of the definitions here}
\begin{lemma}\label{lma:single-correct}
  Starting with $\SinglePredicateInstance$, we will close the goal with $\top$
  if $\exists s \in \Language(\Automaton) \HoldsThat \Map(s) = \MonoidElement$
  and $\bot$ otherwise.
\end{lemma}
\begin{proof}
Since we are always guaranteed to terminate following
\cref{lma:single-terminates}, we know that we will close the goal eventually.
Additionally, the property of \cref{lma:single-terminates} is trivially obtained
from the semantics of the $\Image$ predicate as defined in
\cref{def:single-image}. This means that we only need to prove the preservation
of those properties to prove \cref{lma:single-correct}.

\Fudge{I don't understand.}
  % \[
  % \begin{aligned}
  %   & \inferrule
  %   {\Connected(\Automaton, \Filter) \land \FlowEq(\Automaton, \Filter) \land \MonoidElement = \sum_{\Transition \in \Transitions_\Automaton} \Filter(\Transition) \cdot \Map(\Transition), \SomeInequalities, \SomeClause}
  %   {\SinglePredicateInstance, \SomeInequalities, \SomeClause} &\beta \models R_0 \iff \exists i : \beta \models R_i
  % \end{aligned}
  % \]

\begin{lemma}\label{lma:expand-preserves}
  \Expand{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}

  After application of \Expand{}, $\FlowEq$ guarantees that any
  $\Filter(\Transition)$  for every $\Transition \in \Automaton$ is consistent
  with the flow through the automaton. After its application, the proof goal
  either contains an unsatisfiable set of Presburger formulae if the flow is
  infeasible, otherwise a satisfiable set. An infesaible flow can happen in
  either of these conditions:
  \begin{itemize}
    \item if $\Filter$ demands that some transition $\Transition$ is visited more times than would be allowed by $\Automaton$, or
    \item if the path described by $\Filter$ is not consistent with a path from
    $\InitialState$ to some state $\State$ in $\AcceptingStates$.
  \end{itemize}

  Note that \Expand{} is taken directly from the existing formulation of
  \eqref{eq:generate-parikh}.
\end{proof}

\begin{lemma}\label{lma:split-preserves}
  \Split{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}
  \Split{} allows us to case-split at most once on each transition $\Transition
  \in \Transitions$. This means that our proof tree at its leaves will be
  guaranteed to have all transitions of known bounds in the correct format. This
  means that once \Split{} is at fixpoint, the status (present or absent) of any
  transition is known. Therefore it preserves the semantics of the $\Image$
  predicate.
\end{proof}

\begin{lemma}\label{lma:propagate-preserves}
  \Propagate{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}
  \Propagate{} will lazily enforce connectedness to any state. If it can be
  applied, there exists a cut $C$ in $\Automaton$ such that some state $\State$
  for any incoming transition $\Transition \in \Transitions_\Automaton$ has
  $\Filter(t) = 0$  as part of the equations of the proof goal, and such that
  $\State$ has an outgoing transition $\Transition'$ whose lower bound is not
  $0$. Adding this conclusion will make equations unsatisfiable precisely if
  there exists a transition $\Transition \in \Transitions_\Automaton$ such that
  $\Filter(\Transition)$ is inconsistent with the homomorphic image, since it
  implies that the selection of transitions is disconnected.
\end{proof}

\begin{lemma}\label{lma:subsume-preserves}
  \Subsume{} preserves the semantics of $\Image$.
\end{lemma}
\begin{proof}
  Since this rule can \emph{only} be applied if no other rule can, it means that
  the upper and lower bounds of any transition $\Transition \in \Transitions$ as
  defined by $\Filter(\Transition)$ is known, meaning clauses of the shape
  $\Filter(\Transition) > 0$ or $\Filter(\Transition) = 0$ are present in the
  proof goal for every transition and that the $\Image$ predicate has been
  consumed by \Expand{} to be replaced by $\Connected$. This in turns means that
  the goal can be closed only if the assignment of $\Filter$ is consistent with
  an accepting path through $\Automaton$.

  It also means that \Propagate{} must have been applied whenever possible, meaning that for any transition $\Transition$ starting in a state $\State$ that is disconnected the corresponding $\Filter$ values, we must have applied $\Propagate{}$ to derive $\Filter(\Transition) = 0$. Note that for an infeasible assignment (such as an unsatisfiable instance) this would contradict other decisions, but it would preserve the semantics of $\Image$.
\end{proof}

Since the final step we execute is \Subsume, which we know from
\cref{lma:subsume-preserves} preserves the semantics of $\Image$ and each step
along the way
(\cref{lma:expand-preserves,lma:split-preserves,lma:propagate-preserves}), it
holds that \Calculus{} for one automaton also preserves the semantics.
\end{proof}

>>>>

We write commutative monoids as $\Monoid =
\left(X;+_{\Monoid};0_{\Monoid}\right)$, where we mean $X$ to be the underlying
set of elements, $+_{\Monoid}$ to be the (commutative) monoid map, with
$0_{\Monoid}$ as its neutral element.

@@

We define an alphabet as a set of symbols $\Alphabet$ with words $\Strings$, and
the concatenation operation as $s_1 \Concat{} s_2$ over two strings $s_1, s_2$.

@@

\subsection{The Parikh image of a regular language expressed in Presburger arithmetic}
\label{sec:verma}

@@

It is easy to see that the Parikh map~$\ParikhMap$ represents a
homomorphism from the (free) non-commutative monoid~$\Strings$ to the
(free) commutative monoid~$\Naturals^k$. As we are often not
interested in the full Parikh image, but only in some projections of
it, for the purposes of this paper we use a generalised version of the
Parikh map. We consider \emph{arbitrary} homomorphisms
$\Map:\: \Sigma^* \to \Monoid$, where
$\Monoid = \left(X;+_{\Monoid};0_{\Monoid}\right)$ is a commutative
monoid; this means that $\Map(\epsilon) = 0_{\Monoid}$ and
$\Map(u \Concat w) = \Map(u) +_{\Monoid} \Map(w)$. We give several
examples of such generalised Parikh image later in this section.

@@

\begin{definition}\label{def:single-image}
  The Parikh predicate, $\SinglePredicateInstance$, for some automaton
  $\Automaton = \AutomatonTuple$, modulo some map $\Map$ to a
  commutative monoid $\Monoid$ as described in \cref{sec:generalised}
  and with a transition selection function
  $\Filter:\: \Transitions \to \Naturals$ holds when
  $\MonoidElement \in \Monoid$ is the Parikh image of $\Automaton$
  modulo $\Map$, or more formally when there is an accepting path
  $\Path = \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ such
  that $\Filter(\Transition) = \TransitionCount(\Transition, \Path)$
  for all $t \in \Transitions$, and $\MonoidElement = \Map(\WordOf(\Path))$.
\end{definition}

@@

Note that the filtering function~$\Filter$ is evaluated symbolically in these
rules, and can in practice be read as a function from transitions to
$\Naturals$-valued terms (e.g.~\texttt{t} or~\texttt{t+1}). In our implementation \Catra{},
described in \cref{sec:implementation}, $\Filter$~is a vector of fresh variables
with the same size as~$\Transitions$.

We use the shorthand notation~$\Transitions_\Automaton$ to refer to the
transitions of an automaton~$\Automaton$. Additionally, for an
automaton~$\Automaton = \AutomatonTuple$ we allow mapping the selection function
like so: $\Filter(\Automaton) = \Tuple{\States, \InitialState, \AcceptingStates,
\Set{\Transition \in \Transitions \SuchThat \Filter(\Transition) > 0}}$, i.e.
$\Automaton$~with only the transitions for which~$\Filter$ is positive. In this
instance, the basis for the matched linear inequalities is
implicitly~$\SomeInequalities$. Similarly, for our commutative monoid $\Monoid =
\left(X;+_{\Monoid};0_{\Monoid}\right)$ and the map into it $\Map : \Strings
\rightarrow X$, we also allow mapping over transitions:
$\Map(\FromLabelTo{\State}{\Label}{\State'}) = \Map(\Label)$.

Each of the rules of \cref{tbl:rules:single} should be read and
applied bottom-up, and relates
premises~$\SomeClause_1, \ldots, \SomeClause_k$ with some
conclusion~$\SomeClause$. When constructing a proof, we start from some
root~$\SomeClause$, and then apply proof rules to the goals of the proof
until a goal can be closed, or no more rule is applicable. Since
our rules operate by adding and matching linear (in)equalities in a proof goal,
we use the shorthand of listing the matched inequalities as antecedents. An
example of this can be seen in the~\Propagate{} rule. \Fudge{We pull solutions
out of a hat? I have no idea.}

@@

\subsubsection{The rules in Table~\ref{tbl:rules:single} are solution-preserving}

\begin{lemma}\label{lma:single-correct}
  Consider an application of one of the rules in
  Table~\ref{tbl:rules:single}, with
  premises~$\SomeClause_1, \ldots, \SomeClause_k$ and
  conclusion~$\SomeClause$. An assignment~$\beta$ satisfies the
  conclusion~$\SomeClause$ if and only if it satisfies one of the
  premises~$\SomeClause_i$.
\end{lemma}

\begin{proof}
  This property has to be shown by analysing the possible applications
  of each proof rule.

  \Expand{} unfolds the definition of the $\Image$ predicate. To show
  that the rule is solution-preserving we prove the equivalence of the
  upper and lower sets of formulas:
  \begin{itemize}
  \item Assume that $\beta$ satisfies the conclusion, which means that
    there is some accepting path
    $\Path = \PathEnumeration \in \Accepting{\Paths(\Automaton)}$ with
    $\val_\beta(\Filter(\Transition)) = \TransitionCount(\Transition,
    \Path)$ and $\val_\beta(\MonoidElement) =
    \Map(\WordOf(\Path))$. Since immediately implies that $\beta$
    satisfies $\Connected(\Automaton, \Filter)$, since a path is
    connected, and $\FlowEq(\Automaton, \Filter)$ since an accepting
    path satisfies the flow equations. The
    equation~$\MonoidElement = \sum_{\Transition \in
      \Transitions_\Automaton}\Filter(\Transition) \cdot
    \Map(\Transition)$ holds because of
    $\val_\beta(\Filter(\Transition)) = \TransitionCount(\Transition,
    \Path)$.
  \item Assume that $\beta$ satisfies the premise, which implies that
    $\val_\beta(\Filter)$ describes a consistent, connected flow of
    the automaton. By the same argument as in
    \cite{generate-parikh-image} (Section~\ref{sec:verma}), this flow
    can be mapped to an accepting path~$\Path$ of $\Automaton$ such
    that each transition~$\Transition$ occurs on $\Path$ exactly
    $\val_\beta(\Filter(\Transition))$ times. Together with the equation
    $\val_\beta(\Filter(\Transition)) = \TransitionCount(\Transition,
    \Path)$, this implies that $\beta$ satisfies $\SinglePredicateInstance$.
  \end{itemize}

  In \Split{}, we make use of the fact that $\Filter(\Transition)$ is
  $\Naturals$-valued by definition. For any $\beta$, clearly exactly
  one of $\Filter(\Transition) = 0$ or $\Filter(\Transition) > 0$ will
  be satisfied, implying the property.

  For \Propagate{}, suppose that
  $\SeparatingCut(C, \Automaton, \State)$, which means that every
  accepting path visiting~$q$ contains at least one of the transitions
  in $C$. For a $\beta$ satisfying $\Connected(\Automaton, \Filter)$
  and $\Filter(\Transition') = 0$ for $\Transition' \in C$, this means
  that also $\val_\beta(\Filter(\Transition)) = 0$ has to hold.

  Finally, for \Subsume{}, observe that if \Split{} cannot be applied,
  then a goal must contain~$\Filter(\Transition) = 0$ or
  $\Filter(\Transition) > 0$ for every $\Transition$. In case the
  formulas in $\SomeInequalities$ are inconsistent, an application of
  \Subsume{} is trivially solution-preserving; therefore assume that
  $\SomeInequalities$ is consistent, which means that it contains
  exactly one of $\Filter(\Transition) = 0$ or
  $\Filter(\Transition) > 0$ for each $\Transition$.
\end{proof}
